---
title: "QC pipeline for mass spetrometry data"
author: "Chacha Issarow"
date: "2025-06-20"
  output: html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Overview of the pipeline:
#Quality control (QC) pipeline for both DIA and DDA mass spectrometry data using R packages. 
#The pipeline includes:
# Data and annotation importation files
# Data filtering and normalization
# QC plots (missing values, intensity distributions, data completeness, PCA, CVs, data correlation, heatmap, etc)
 
#This pipeline provides an overview of how you can prepare the quantitative protein/peptide matrix output from common search engines and software, such as Spectronaut, MaxQuant, Proteome Discoverer, Skyline and Dia-nn.

## R Markdown (This pipeline was written in R markdown. You can run it as R script by removing R markdown conditions (```{r} and ```) from Load packages)
#Load packages
```{r}
library(protti)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(tidyr)
library(readr)
library(pheatmap)
library(tibble)
```

#Include the path and import the data and annotation here:
```{r, warning=FALSE}
setwd("E:/TB-HART_allData/DIA-NN_secondResults") # Replace with your actual file paths

data <- read_delim("report.pg_matrix.tsv", show_col_types = FALSE) # DIA or DDA data 
annotation <- read_csv("annotationEdit.csv", show_col_types = FALSE) # Annotation file, contains sample meta info (e.g. sample, condition, batch)
dim(data)
head(annotation)
```

#Convert wide file (like report.pg_matrix.csv from Dia-nn) to long format. If using long file, such as report.tsv from Dia-nn, then skip this stage.
#Convert wide file (like report.gg_matrix.csv) to long format
```{r}
data <- data %>% pivot_longer(cols = starts_with("2"),
                              names_to = "sample",
                              values_to = "intensity"
                                )
dim(data)
```

#Merge data with annotation file
```{r}
data_annotated <- data %>% left_join(y = annotation, by = "sample")
dim(data_annotated)
```

#Transform intensities to Log2 
```{r}
data_annotated <- data_annotated %>%
  mutate(intensity_log2 = log2(intensity))
names(data_annotated)
```


#Filter only samples prior normalisation, remove pools, PPD and NA 
#use this option if you want to remove some samples based on condition, otherwise, just skip it.
```{r}
filtered_sample <- subset(data_annotated, Condition != "Pool")
filtered_sample <- subset(filtered_sample, Condition != "PPD")
filtered_sample <- subset(filtered_sample, Condition != "Baseline_NA")
filtered_sample <- subset(filtered_sample, Condition != "Rechallenge_NA")
dim(filtered_sample)
#names(filtered_sample)
```

#Optional: peptide/protein filtering: remove low quality data.
#filter out proteins with low identification counts or missing intensities.
```{r}
filtered_sample <- filtered_sample %>%  
  group_by(Condition, Protein.Group) %>%
  summarise(
    n = n(),
    n_missing = sum(is.na(intensity_log2)),
    missingness = n_missing / n,
    .groups = "drop"
  ) %>%
  filter(missingness <= 0.7) %>%
  dplyr::select(Protein.Group) %>%
  distinct() %>%
  inner_join(filtered_sample, by = "Protein.Group")
dim(filtered_sample )
```


#Data normalization (e.g., median). Only apply normalzsation if data were not normalised in upstream step (using Dia-NN, Spectronaught, etc). 
#Otherwise, skip this step if data has been normlised in upstream step
```{r}
data_normalised <- filtered_sample %>%
  normalise(
    sample = sample,
    intensity_log2 = intensity_log2,
    method = "median"
  )
#names(data_normalised)
```


#QC assessment
#checking CV
```{r, warning=FALSE}
qc_cvs(
  data = data_annotated,
  grouping = Protein.Group,  #Protein.Ids , #rotein_id
  condition = Condition,
  intensity = intensity,
  plot = TRUE,
  plot_style = "violin"
)
```

#CV in table table format
```{r}
cv <- qc_cvs(
  data = data_annotated,
  grouping = Protein.Group,  #protein_ids, #peptide,
  condition = Condition,
  intensity = intensity,
  plot = FALSE
)
cv
```

#Protein identifications per sample
```{r, fig.width=16, fig.height=10}
#(groups use protein, but precursor_id use peptides)
qc_ids(data = data_annotated,
       sample = sample,
       grouping = Protein.Group, 
       intensity = intensity_log2,
       condition = Condition,
       title = "Protein identifications per sample",
       plot = TRUE
       )
```

Protein identifications per sample in table format
```{r}
qc_ids(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group, 
  intensity = intensity_log2,
  condition = Condition,
  plot = FALSE
)
```

#Sample intensities:
#Intensity boxplot is helpful for quick assessment of any major sample losses or measurement issues.
```{r, warning=FALSE, fig.width=16, fig.height=10}
#Intensity boxplot prior normalisation
ggplot(filtered_sample, aes(x = sample, y = intensity_log2, fill = Condition)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Log2 Intensity per Sample", y = "Log2 Intensity") +
  theme(axis.text.x = element_text(angle = 80, hjust = 1)) +theme(text=element_text(size=16))
```


```{r, warning=FALSE, fig.width=16, fig.height=10}
#Boxplot post normalisation
ggplot(data_normalised, aes(x = sample, y = normalised_intensity_log2, fill = Condition, group_by = Condition)) +
  geom_boxplot() + theme_bw() + theme(text = element_text(size = 16)) +
  theme(axis.text.x = element_text(angle = 80, hjust = 1)) +
  labs(x = "", y = "Intensity", title = "Sample intensities")  #+ theme(legend.position = "none")

```

#Intensities can also be assessed by plotting the median as a line plot.This helps you quickly assess if there are any trends in your data
```{r, fig.width=16, fig.height=10}
qc_median_intensities(
  data = filtered_sample,
  sample = sample,
  grouping = Protein.Group, #protein_ids, #peptide,
  intensity = intensity_log2
)
```

#Data completeness:
#Data completeness checks how many of all detected precursors, peptides or proteins were identified in each sample. 
#The function can return either a plot or a table.
```{r, fig.width=16, fig.height=10}
qc_data_completeness(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  plot = TRUE
)
```

#Or display data completeness in table format
```{r}
comp <- qc_data_completeness(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  plot = FALSE
)
comp
```


#Log2 Intensity distribution pre- or prio-normalisation
#For different kinds of analyses (e.g. t-tests) it is important that your data intensity follows a normal distribution
```{r}
qc_intensity_distribution(
  data = data_annotated,
  grouping = Protein.Group, #protein_group,
  intensity_log2 = intensity_log2,
  plot_style = "histogram"
)
```

#Sample correlation using pheatmap
```{r}
wide_data <- filtered_sample %>%
  dplyr::select(Protein.Group, sample, intensity_log2) %>%
  pivot_wider(names_from = sample, values_from = intensity_log2)
#head(wide_data)
```

```{r}
# Remove the protein_id column for correlation
cor_matrix <- cor(wide_data[,-1], use = "pairwise.complete.obs", method = "pearson") #spearman, pearson 
#head(cor_matrix)
```


```{r, fig.width=16, fig.height=10}
pheat <- pheatmap(cor_matrix,
         main = "Sample Correlation Heatmap",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 10,
         fontsize_row = 10,
         fontsize_col = 10,
         color = colorRampPalette(c("blue", "white", "red"))(100))
pheat
#ggsave("E:/TB-HART_allData/DIA-NN_secondResults/corr_heatmap.pdf", plot = pheat, width = 8, height = 6, dpi = 300)
```


#Principal component analysis (PCA)
#Before plotting your PCA you can check how much of your variance can be explained with each of the principle components
```{r, fig.width=16, fig.height=10}
qc_pca(
  data = filtered_sample, #use "data_normalised" and "normalised_intensity_log2" if applied normalisation in this pipeline
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  condition = Condition,
  digestion = NULL,
  plot_style = "scree"
)
```

#Principal component analysis (PCA)
#This helps us to quickly assess how similar or different our replicates and conditions are
```{r}
#Identify variance and use ggplot
pc <- qc_pca(
  data = filtered_sample,  #use "data_normalised" and "normalised_intensity_log2" if applied normalisation in this pipeline
  sample = sample,
  grouping = Protein.Group, #precursor_id, #protein_group,
  intensity = intensity_log2,
  condition = Condition,
  components = c("PC1", "PC2"),
  plot_style = "pca",
  digestion = NULL
)
#head(pc)
```

```{r}
ggplot(pc$data, aes(x = PC1, y = PC2, colour = Condition)) +
  geom_point(size = 4) + #stat_ellipse(type = "norm", linetype = 2) +
  #stat_ellipse(type = "t") +
  theme_bw() + theme(text = element_text(size = 20)) + #stat_ellipse() +
  #theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "PC1",y = "PC2", title = "Principal component analysis")  #+ stat_ellipse(geom="polygon", level=0.95, alpha=0.2)
#+ theme(legend.position = "none")
#ggsave("E:/TB-HART_allData/DIA-NN_secondResults/pca.pdf", width = 8, height = 6, dpi = 300)
```


##Ranked intensity distribution
# Plot ranked peptide intensities
##It can be useful to check which precursors, peptides or proteins are the most or least abundant in the sample by ranking their intensities
```{r, warning=FALSE}
qc_ranked_intensities(
  data = filtered_sample,
  sample = sample,
  grouping = Protein.Group,
  intensity_log2 = intensity_log2,
  plot = TRUE
)

````


Plotting a heatmap using pheatmap
```{r}
#reshape data to wide format
heatmap_matrix <- data_normalised %>%
  drop_na(intensity_log2) %>%
  dplyr::select(Protein.Group, sample, intensity_log2) %>%
  pivot_wider(names_from = sample, values_from = intensity_log2) %>%
  column_to_rownames("Protein.Group") %>%
  as.matrix()
#head(heatmap_matrix)
```

```{r}
# Top 50 or 100 most variable proteins
#top_var <- apply(heatmap_matrix, 1, var) %>% sort(decreasing = TRUE)
#heatmap_matrix <- heatmap_matrix[names(top_var)[1:20], ]
#head(heatmap_matrix)
```


```{r}
# Remove rows with NA/NaN/Inf
heatmap_matrix <- heatmap_matrix[complete.cases(heatmap_matrix), ]
#head(heatmap_matrix)
```


```{r}
# Prepare annotation for columns
sample_annotation <- annotation %>%
  column_to_rownames("sample")
#head(sample_annotation)
```

```{r}
all(colnames(heatmap_matrix) %in% rownames(sample_annotation))  # should be TRUE
#all(rownames(sample_annotation) %in% colnames(heatmap_matrix))  # should also be TRUE
```

```{r}
#Plot Heatmap with pheatmap
pheat <- pheatmap(
  heatmap_matrix,
  annotation_col = sample_annotation,
  scale = "row", # optional: scales each row (z-score)
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = FALSE,
  show_colnames = FALSE
)
ggsave("E:/TB-HART_allData/DIA-NN_secondResults/pheatmap_heatmap2.pdf", plot = pheat, width = 8, height = 6, dpi = 300) #Replace with your actual path
pheat
```


