---
title: "QC, downstream and gene ontology analysis pipeline for mass spectrometry data"
author: "Chacha Issarow"
date: "2025-07-10"
  output: html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#This pipeline for QC, downstream analysis and gene ontology for mass spectrometry data (DIA and DDA), was written in R markdown. 
#You can also copy from packages (here) downwards to a normal R script and run it if R markdown is not installed in your machine. 
#If not uing R markdown, please remember to remove R markdown conditions (e,g, ```{r} and ```) before running.
## R Markdown
# Load packages
```{r}
library(protti)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(tidyr)
library(readr)
library(pheatmap)
library(tibble)
library(limma)
library(org.Hs.eg.db)  # Use appropriate organism package (e.g., org.Mm.eg.db for mouse)
library(clusterProfiler)
library(enrichplot)
library(DOSE)
library(pathview)
#library(tidyverse)
```

#Indlude the path and import the data here:
```{r, warning=FALSE}
setwd("E:/TB-HART_allData/DIA-NN_secondResults") # Replace with your actual file paths
data <- read_delim("report.pg_matrix.tsv", show_col_types = FALSE) # DIA or DDA data 
annotation <- read_csv("annotationEdit.csv", show_col_types = FALSE) # Contains sample meta info (e.g. sample, condition, batch)
dim(data)
#dim(annotation)
```

#Convert wide file (like report.gg_matrix.csv) to long format
```{r}
data <- data %>% pivot_longer(cols = starts_with("2"),
                              names_to = "sample",
                              values_to = "intensity"
                                )
dim(data)
```

#Merge data with annotation file
```{r}
data_annotated <- data %>% left_join(y = annotation, by = "sample")
dim(data_annotated)
```

#Transform intensities to Log2 
```{r}
data_annotated <- data_annotated %>%
  mutate(intensity_log2 = log2(intensity))
names(data_annotated)
```

#Filter only samples prior normalisation, remove pools, PPD and NA 
#use  this option if you want to remove some samples based on conditions, if not just skip it.
```{r}
filtered_sample <- subset(data_annotated, Condition != "Pool")
filtered_sample <- subset(filtered_sample, Condition != "PPD")
filtered_sample <- subset(filtered_sample, Condition != "Baseline_NA")
filtered_sample <- subset(filtered_sample, Condition != "Rechallenge_NA")
dim(filtered_sample)
names(filtered_sample)
```

#Optional: peptide/protein filtering: remove low quality data.
#filter out proteins with low identification counts or missing intensities.
```{r}
filtered_sample <- filtered_sample %>%
  group_by(Condition, Protein.Group) %>%
  summarise(
    n = n(),
    n_missing = sum(is.na(intensity_log2)),
    missingness = n_missing / n,
    .groups = "drop"
  ) %>%
  filter(missingness <= 0.7) %>%
  dplyr::select(Protein.Group) %>%
  distinct() %>%
  inner_join(filtered_sample, by = "Protein.Group")
dim(filtered_sample )
```


#Data normalization (e.g., median). Only apply normalzsation if data were not normalised in upstream step (using Dia-NN, Spectronaught, etc). 
#Otherwise, skip this step if data has been normlised in upstream step
```{r}
data_normalised <- filtered_sample %>%
  normalise(
    sample = sample,
    intensity_log2 = intensity_log2,
    method = "median"
  )
#names(data_normalised)
```


#QC assessment
#checking CV
```{r, warning=FALSE}
qc_cvs(
  data = data_annotated,
  grouping = Protein.Group,  #Protein.Ids , #rotein_id
  condition = Condition,
  intensity = intensity,
  plot = TRUE,
  plot_style = "violin"
)
```

#CV in table table format
```{r}
cv <- qc_cvs(
  data = data_annotated,
  grouping = Protein.Group,  #protein_ids, #peptide,
  condition = Condition,
  intensity = intensity,
  plot = FALSE
)
cv
```

#Protein identifications per sample
```{r, fig.width=16, fig.height=10}
#(groups use protein, but precursor_id use peptides)
qc_ids(data = data_annotated,
       sample = sample,
       grouping = Protein.Group, 
       intensity = intensity_log2,
       condition = Condition,
       title = "Protein identifications per sample",
       plot = TRUE
       )
```

Protein identifications per sample in table format
```{r}
qc_ids(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group, 
  intensity = intensity_log2,
  condition = Condition,
  plot = FALSE
)
```

#Sample intensities:
#Intensity boxplot is helpful for quick assessment of any major sample losses or measurement issues.
```{r, warning=FALSE, fig.width=16, fig.height=10}
#Intensity boxplot prior normalisation
ggplot(filtered_sample, aes(x = sample, y = intensity_log2, fill = Condition)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Log2 Intensity per Sample", y = "Log2 Intensity") +
  theme(axis.text.x = element_text(angle = 80, hjust = 1)) +theme(text=element_text(size=16))
```


```{r, warning=FALSE, fig.width=16, fig.height=10}
#Boxplot post normalisation
ggplot(data_normalised, aes(x = sample, y = normalised_intensity_log2, fill = Condition, group_by = Condition)) +
  geom_boxplot() + theme_bw() + theme(text = element_text(size = 16)) +
  theme(axis.text.x = element_text(angle = 80, hjust = 1)) +
  labs(x = "", y = "Intensity", title = "Sample intensities")  #+ theme(legend.position = "none")

```

#Intensities can also be assessed by plotting the median as a line plot.This helps you quickly assess if there are any trends in your data
```{r, fig.width=16, fig.height=10}
qc_median_intensities(
  data = filtered_sample,
  sample = sample,
  grouping = Protein.Group, #protein_ids, #peptide,
  intensity = intensity_log2
)
```

#Data completeness:
#Data completeness checks how many of all detected precursors, peptides or proteins were identified in each sample. 
#The function can return either a plot or a table.
```{r, fig.width=16, fig.height=10}
qc_data_completeness(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  plot = TRUE
)
```

#Or display data completeness in table format
```{r}
comp <- qc_data_completeness(
  data = data_annotated,
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  plot = FALSE
)
comp
```


#Log2 Intensity distribution pre- or prio-normalisation
#For different kinds of analyses (e.g. t-tests) it is important that your data intensity follows a normal distribution
```{r}
qc_intensity_distribution(
  data = data_annotated,
  grouping = Protein.Group, #protein_group,
  intensity_log2 = intensity_log2,
  plot_style = "histogram"
)
```

#Sample correlation using pheatmap
```{r}
wide_data <- filtered_sample %>%
  dplyr::select(Protein.Group, sample, intensity_log2) %>%
  pivot_wider(names_from = sample, values_from = intensity_log2)
#head(wide_data)
```

```{r}
# Remove the protein_id column for correlation
cor_matrix <- cor(wide_data[,-1], use = "pairwise.complete.obs", method = "pearson") #spearman, pearson 
#head(cor_matrix)
```


```{r, fig.width=16, fig.height=10}
pheat <- pheatmap(cor_matrix,
         main = "Sample Correlation Heatmap",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 10,
         fontsize_row = 10,
         fontsize_col = 10,
         color = colorRampPalette(c("blue", "white", "red"))(100))
pheat
#ggsave("E:/TB-HART_allData/DIA-NN_secondResults/corr_heatmap.pdf", plot = pheat, width = 8, height = 6, dpi = 300)
```


#Principal component analysis (PCA)
#Before plotting your PCA you can check how much of your variance can be explained with each of the principle components
```{r, fig.width=16, fig.height=10}
qc_pca(
  data = filtered_sample, #use "data_normalised" and "normalised_intensity_log2" if applied normalisation in this pipeline
  sample = sample,
  grouping = Protein.Group,
  intensity = intensity_log2,
  condition = Condition,
  digestion = NULL,
  plot_style = "scree"
)
```

#Principal component analysis (PCA)
#This helps us to quickly assess how similar or different our replicates and conditions are
```{r}
#Identify variance and use ggplot
pc <- qc_pca(
  data = filtered_sample,  #use "data_normalised" and "normalised_intensity_log2" if applied normalisation in this pipeline
  sample = sample,
  grouping = Protein.Group, #precursor_id, #protein_group,
  intensity = intensity_log2,
  condition = Condition,
  components = c("PC1", "PC2"),
  plot_style = "pca",
  digestion = NULL
)
#head(pc)
```

```{r}
ggplot(pc$data, aes(x = PC1, y = PC2, colour = Condition)) +
  geom_point(size = 4) + #stat_ellipse(type = "norm", linetype = 2) +
  #stat_ellipse(type = "t") +
  theme_bw() + theme(text = element_text(size = 20)) + #stat_ellipse() +
  #theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "PC1",y = "PC2", title = "Principal component analysis")  #+ stat_ellipse(geom="polygon", level=0.95, alpha=0.2)
#+ theme(legend.position = "none")
#ggsave("E:/TB-HART_allData/DIA-NN_secondResults/pca.pdf", width = 8, height = 6, dpi = 300)
```


##Ranked intensity distribution
# Plot ranked peptide intensities
##It can be useful to check which precursors, peptides or proteins are the most or least abundant in the sample by ranking their intensities
```{r, warning=FALSE}
qc_ranked_intensities(
  data = filtered_sample,
  sample = sample,
  grouping = Protein.Group,
  intensity_log2 = intensity_log2,
  plot = TRUE
)

```


##Downstream analysis
#Plotting a heatmap using pheatmap
```{r}
#reshape data to wide format
heatmap_matrix <- filtered_sample %>%
  drop_na(intensity_log2) %>%
  dplyr::select(Protein.Group, sample, intensity_log2) %>%
  pivot_wider(names_from = sample, values_from = intensity_log2) %>%
  column_to_rownames("Protein.Group") %>%
  as.matrix()
#head(heatmap_matrix)
```

#Use this step if you want to filter top proteins, such as 100 or 50 top variableproteins
```{r}
# Use this for top 50 or 100 most variable proteins
top_var <- apply(heatmap_matrix, 1, var) %>% sort(decreasing = TRUE)
heatmap_matrix <- heatmap_matrix[names(top_var)[1:20], ]
#head(heatmap_matrix)
```


```{r}
# Remove rows with NA/NaN/Inf
heatmap_matrix <- heatmap_matrix[complete.cases(heatmap_matrix), ]
#head(heatmap_matrix)
```


```{r}
# Prepare annotation for columns
sample_annotation <- annotation %>%
  column_to_rownames("sample")
#head(sample_annotation)
```

```{r}
all(colnames(heatmap_matrix) %in% rownames(sample_annotation))  # should be TRUE
```

```{r, fig.width=12, fig.height=8}
#Plot Heatmap with pheatmap
pheat <- pheatmap(
  heatmap_matrix,
  annotation_col = sample_annotation,
  scale = "row", # optional: scales each row (z-score)
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = FALSE,
  show_colnames = FALSE
  #fontsize = 10, #col and row fontsize
  #fontsize.legend = 7  #Legend fontsize
)
#ggsave("E:/TB-HART_allData/DIA-NN_secondResults/pheatmap_heatmap.pdf", plot = pheat, width = 12, height = 8) #Cange this to your own pth
pheat
```

 
##Differential expression using limma
```{r}
annotation <- annotation %>% filter(Condition != "Pool") %>% 
  filter(Condition != "PPD") %>% filter(Condition != "Baseline_NA") %>% filter(Condition != "Rechallenge_NA")
dim(annotation)
```
 
 
```{r}
# Prepare for limma
design <- model.matrix(~ 0 + annotation$Condition)
colnames(design) <- levels(as.factor(annotation$Condition))
rownames(design) <- annotation$sample
dim(design)
```

```{r}
# Create ExpressionSet-like object
intensity_matrix <- filtered_sample  %>% #use normlised data if applicable 
  dplyr::select(Protein.Group, sample, intensity_log2) %>% 
  tidyr::pivot_wider(names_from = sample, values_from = intensity_log2) %>%
  column_to_rownames("Protein.Group") %>% as.matrix()
dim(intensity_matrix)
```


```{r}
#Match order
intensity_matrix <- intensity_matrix[, rownames(design)]  # match column order
dim(intensity_matrix)
```

```{r}
all(colnames(intensity_matrix) == rownames(design))  # Should be TRUE
```


```{r, warning=FALSE}
fit <- lmFit(intensity_matrix, design)
#head(fit)
```


```{r}
# Define contrasts
contrast_matrix <- makeContrasts(
  Baseline_RTB_vs_Rechallenge_RTB = Baseline_RTB - Rechallenge_RTB,
  levels = design
)
#head(contrast_matrix)
```


```{r}
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
#head(fit2)
```

```{r}
# Get DE results
results <- topTable(fit2, coef = "Baseline_RTB_vs_Rechallenge_RTB", adjust = "BH", number = Inf)
#head(results)
```

#Add proein names and export
```{r}
results$Protein.Group <- rownames(results)
results <- results %>%
  relocate(Protein.Group)
#head(results)

#write.csv(results, "limma_DE_results.csv", row.names = FALSE)
```

#Volcano plots
```{r}
results$Diff_expression <- "Not significant"
results$Diff_expression[results$logFC > 1 & results$P.Value < 0.05] <- "Up regulated"
results$Diff_expression[results$logFC < -1 & results$P.Value < 0.05] <- "Down regulated"
mycolors <- c("grey", "red", "blue")
names(mycolors) <- c("Not significant", "Up regulated", "Down regulated")

#head(results)
```

```{r, warning=FALSE}
ggplot(results, aes(x = logFC, y=-log10(P.Value), color = Diff_expression)) +
  geom_point(size =4) + theme_bw() + xlim(-4, 4) +
  #scale_colour_discrete(type=ghibli_subset, breaks = c("Up", "Down")) +
  geom_hline(yintercept = -1*log10(0.05), linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = -1, linetype = 2, alpha = 0.5) + #xlim(-2, 2) +
  scale_colour_manual(values = mycolors) + theme(text = element_text(size = 18)) +
  labs(x = "log2(fold change)", y = "-log10(p-value)", title =  "Baseline_RTB_vs_Rechallenge_RTB")
```

Gene ontology enrichment using ClusreProfiler. Use DE results from limma as input for ClusterProfiler.
```{r}
gene_ids <- results$Protein.Group
#head(gene_ids)
```

```{r}
# Convert to Entrez IDs (if needed)
gene_df <- bitr(gene_ids, fromType = "UNIPROT", 
                toType = "ENTREZID", 
                OrgDb = org.Hs.eg.db)
#head(gene_df)
```

```{r}
gene_ids <- gene_df$ENTREZID
#head(gene_ids)
```

```{r}
##GO Enrichment Analysis
ego <- enrichGO(gene = gene_ids,
                OrgDb = org.Hs.eg.db,
                ont = "BP",  # Use "MF" or "CC" for other ontologies
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.2,
                readable      = TRUE)
#head(ego)
```

```{r}
#Visualisation
# Barplot
barplot(ego, showCategory = 10, title = "GO Enrichment")
```

```{r}
# Dotplot
dotplot(ego, showCategory = 10) + 
  ggtitle("Biological Process enrichment")
```

```{r, warning=FALSE, fig.width=16, fig.height=10}
# cnetplot - gene-concept network
cnetplot(ego, circular = TRUE, colorEdge = TRUE)
```

#KEGG Pathway Enrichment
```{r}
ekegg <- enrichKEGG(gene = gene_ids, 
                    organism = 'hsa',  # use 'mmu' for mouse
                    pvalueCutoff = 0.05)
#head(ekegg)
```


```{r}
# Optional: convert to readable symbols
ekegg <- setReadable(ekegg, 
                     OrgDb = org.Hs.eg.db, 
                     keyType="ENTREZID")
#head(ekegg)
```

```{r}
##Visulisation
# KEGG Dotplot
dotplot(ekegg, showCategory = 10)
```


```{r, warning=FALSE, fig.width=16, fig.height=10}
# KEGG Pathway Plot (requires pathview), not very clear how to interpret the figure
pathview(gene.data  = gene_ids,
         pathway.id = ekegg@result$ID[1], 
         species    = "hsa")
```


```{r}
#Save results
# Save GO results
write.csv(as.data.frame(ego), "GO_BP_enrichment.csv", row.names = FALSE)

# Save KEGG results
write.csv(as.data.frame(ekegg), "KEGG_enrichment.csv", row.names = FALSE)
```


#Part2: Use fold change (logFC) to separate down and up regulation
```{r}
# we want the log2 fold change 
original_gene_list <- results$logFC
#head(original_gene_list)
```

```{r}
# name the vector
names(original_gene_list) <- results$Protein.Group
#head(original_gene_list)
```

```{r}
# omit any NA values 
gene_list <- na.omit(original_gene_list)
# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)
#head(gene_list)
```


```{r}
##Gene Set Enrichment
organism <- org.Hs.eg.db
gse <- gseGO(geneList=gene_list, 
             ont ="MF",       ##ont one of “BP”, “MF”, “CC” or “ALL”
             keyType = "UNIPROT", 
             #nPerm = 10000, 
             minGSSize = 10, # exclude GO terms with <10 genes
             maxGSSize = 500, # filter out GO terms with >800 genes
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "none")   #BH
#head(gse)
```


```{r}
###Dotplot
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign) + 
  ggtitle("Significant Biological Process enrichment")
```


```{r}

# categorySize can be either 'pvalue' or 'geneNum'
cnetplot(gse, categorySize="pvalue", foldChange=gene_list, showCategory = 3)

```

#KEGG Gene Set Enrichment Analysis
For KEGG pathway enrichment using the gseKEGG() function, we need to convert id types. We can use the bitr function for this (included in clusterProfiler).

```{r}
#Prepare Input
# Convert gene IDs for gseKEGG function
# We will lose some genes here because not all IDs will be converted
ids<-bitr(names(original_gene_list), fromType = "UNIPROT", toType = "ENTREZID", OrgDb=organism)
#head(ids) 
```

```{r}
# remove duplicate IDS (here I use "ENSEMBL", but it should be whatever was selected as keyType)
dedup_ids <- ids[!duplicated(ids[c("UNIPROT")]),]
#head(dedup_ids)
```

```{r}
# Create a new dataframe df2 which has only the genes which were successfully mapped using the bitr function above
df2 = results[results$Protein.Group %in% dedup_ids$UNIPROT,]
#head(df2)
```

```{r}
# Create a new column in df2 with the corresponding ENTREZ IDs
df2$ENTREZ_ID = dedup_ids$ENTREZID
#head(df2)
```

```{r}
# Create a vector of the gene unuiverse
kegg_gene_list <- df2$logFC
#head(kegg_gene_list)
```

```{r}
# Name vector with ENTREZ ids
names(kegg_gene_list) <- df2$ENTREZ_ID
# omit any NA values 
kegg_gene_list <- na.omit(kegg_gene_list)
# sort the list in decreasing order (required for clusterProfiler)
kegg_gene_list = sort(kegg_gene_list, decreasing = TRUE)
#head(kegg_gene_list)
```

```{r}
any(duplicated(names(kegg_gene_list)))
```

```{r}
#remove duplicates
kegg_gene_list <- kegg_gene_list[!duplicated(names(kegg_gene_list))]
```


```{r}
##Create gseKEGG object
kegg_organism = "hsa"  #
kk2 <- gseKEGG(geneList     = kegg_gene_list,
               organism     = kegg_organism,
               minGSSize    = 3,
               maxGSSize    = 500,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "kegg")
#head(kk2)
```

```{r}
##Dotplot
dotplot(kk2, showCategory = 10, title = "KEGG Enriched Pathways" , split=".sign") + facet_grid(.~.sign)
```

```{r}

##Category Netplot:
#The cnetplot depicts the linkages of genes and biological concepts (e.g. 
#GO terms or KEGG pathways) as a network (helpful to see which genes are involved in 
#enriched pathways and genes that may belong to multiple annotation categories).
# categorySize can be either 'pvalue' or 'geneNum'
cnetplot(kk2, categorySize="pvalue", foldChange=gene_list)
```



